---
title: "Create recommended data splits"
output: html_notebook
---

```{r}
library(tidyverse)
```

# Load metadata

Create metadata file for this repo (do not run)

```{r eval=TRUE}
cpjump1 <- 
  arrow::read_parquet("../../pilot-cpjump1-data/collated/2020_11_04_CPJUMP1/2020_11_04_CPJUMP1_all_normalized_feature_select_outlier_trimmed.parquet", col_select = matches("Metadata")) %>% 
  filter(Metadata_experiment_condition == "Standard") %>%
  mutate(Metadata_timepoint = as.integer(Metadata_timepoint))

timepoint_code <- 
  cpjump1 %>%
  distinct(Metadata_experiment_type, Metadata_timepoint) %>%
  arrange(Metadata_experiment_type, Metadata_timepoint) %>%
  group_by(Metadata_experiment_type) %>%
  mutate(Metadata_timepoint_code = rank(Metadata_timepoint)) %>%
  mutate(Metadata_timepoint_code = factor(Metadata_timepoint_code, labels = c("low", "high")))
  
set.seed(42)

gene_target_random_code <- 
  cpjump1 %>%
  distinct(Metadata_target) %>%
  slice_sample(prop = 1) %>%
  mutate(Metadata_target_radix = seq_along(Metadata_target))

target_across <-
  cpjump1 %>%
  distinct(Metadata_experiment_type, Metadata_target) %>%
  count(Metadata_target) %>% 
  mutate(Metadata_target_is_across = n==3) 

cpjump1 <-
  cpjump1 %>% 
  inner_join(timepoint_code) %>%
  inner_join(gene_target_random_code) %>%
  inner_join(target_across)

cpjump1 %>%
  write_csv("cpjump1_metadata.csv")
```

Load metadata file for CPJUMP1

```{r}
cpjump1 <- vroom::vroom("cpjump1_metadata.csv", show_col_types = FALSE)
```

# Inspect data

Number of wells

```{r}
cpjump1 %>%
  count()
```

Number of plates

```{r}
cpjump1 %>%
  distinct(Metadata_Plate) %>%
  count()
```
Plates per experimental configuration

```{r rows.print=20}
cpjump1 %>%
  distinct(Metadata_cell_line, Metadata_experiment_type, Metadata_timepoint_code, Metadata_timepoint, Metadata_Plate) %>%
  count(Metadata_cell_line, Metadata_experiment_type, Metadata_timepoint, Metadata_timepoint_code)
```
Unique perturbations per experimental configuration

```{r rows.print=20}
cpjump1 %>%
  distinct(Metadata_cell_line, Metadata_experiment_type, Metadata_timepoint_code, Metadata_timepoint, Metadata_broad_sample) %>%
  count(Metadata_cell_line, Metadata_experiment_type, Metadata_timepoint, Metadata_timepoint_code)
```

Genes represented in each perturbation type

`n_gene`s have `n_pert`s targeting/overexpressing them, in that `Metadata_experiment_type`.

```{r}
cpjump1 %>%
  distinct(Metadata_experiment_type, Metadata_broad_sample, Metadata_target) %>%
  count(Metadata_experiment_type, Metadata_target, name = "n_pert") %>%
  count(Metadata_experiment_type, n_pert, name = "n_gene") %>%
  select(Metadata_experiment_type, n_gene, n_pert)
```

Count genes that are represented across all 3 perturbations (`Metadata_target_is_across` is `TRUE`)

```{r}
cpjump1 %>%
  distinct(Metadata_experiment_type, Metadata_target, Metadata_target_is_across) %>%
  count(Metadata_experiment_type, Metadata_target_is_across)
```

```{r}
cpjump1_samples <- 
  cpjump1 %>%
  distinct(Metadata_cell_line, 
           Metadata_experiment_type, 
           Metadata_timepoint_code, 
           Metadata_timepoint, 
           Metadata_target, 
           Metadata_target_is_across,
           Metadata_target_radix,
           Metadata_broad_sample)
```

# Define splits

## Representation learning

Strategy: Per cell line, train on CRISPRs and ORFs, validate on Compounds (one time point), and test on Compounds (the other time point).

Splits are saved at [`representation_learning_split_example_1.csv`](representation_learning_split_example_1.csv).

```{r}
split_sets <- 
  cpjump1_samples %>%
  group_by(Metadata_cell_line) %>%
  mutate(Metadata_split = 
           case_when(
             Metadata_experiment_type %in% c("CRISPR", "ORF") ~ "train",
             Metadata_experiment_type == "Compound" & Metadata_timepoint_code == "low" ~ "validate",
             Metadata_experiment_type == "Compound" & Metadata_timepoint_code == "high" ~ "test"
           )) %>%
  mutate(Metadata_split = as.factor(Metadata_split)) %>%
  mutate(Metadata_split = fct_relevel(Metadata_split, c("train", "validate", "test")))
```


```{r}
split_sets %>%
  count(Metadata_cell_line, Metadata_split) %>%
  group_by(Metadata_cell_line) %>%
  mutate(frac = round(n/sum(n), 2))
```


```{r}
split_sets %>% 
  select(Metadata_split, everything()) %>%
  write_csv("representation_learning_split_example_1.csv")
```

## Gene-compound matching

Strategy: Per cell line and time-point level (i.e. low or high), split by gene target into 60-20-20 for train-validate-test.

Splits are saved at [`gene_compound_split_example_1.csv`](gene_compound_split_example_1.csv).

```{r}
gene_target_split_sets <- 
  cpjump1_samples %>%
  distinct(Metadata_target, Metadata_target_radix, Metadata_target_is_across) %>%
  filter(Metadata_target_is_across)
  
n_genes <- nrow(gene_target_split_sets)
n_train  <- floor(n_genes * .60)
n_validate <- ceiling(n_genes * .20)
n_test <- n_genes - n_train - n_validate

gene_target_split_sets <-
  gene_target_split_sets %>%
  mutate(Metadata_split =
           c(
             rep("train", n_train),
             rep("validate", n_validate),
             rep("test", n_test)
           )) %>%
  mutate(Metadata_split = as.factor(Metadata_split)) %>%
  mutate(Metadata_split = fct_relevel(Metadata_split, c("train", "validate", "test")))

gene_target_split_sets %>%
  count(Metadata_split) %>%
  mutate(frac = round(n/sum(n), 2))
```


```{r}
split_sets <- 
  cpjump1_samples %>%
  inner_join(gene_target_split_sets) %>% 
  filter(Metadata_target_is_across) %>% 
  distinct(Metadata_broad_sample, 
           Metadata_experiment_type, 
           Metadata_target,
           Metadata_split) %>% 
  pivot_wider(values_from = "Metadata_broad_sample", 
              names_from = "Metadata_experiment_type", 
              values_fn = list) %>% 
  unnest(c(ORF, CRISPR, Compound))

split_sets %>%
  count(Metadata_split) %>%
  mutate(frac = round(n/sum(n), 2))
```


```{r}
split_sets %>% 
  select(Metadata_split, everything()) %>%
  write_csv("gene_compound_split_example_1.csv")
```


